Relationships in the extremes and their influence on competition and coexistence

### 2022-01-10 paper/Paper.Rnw
edits to some of the G4 material

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Descamps_2005, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
%***DAN: Jasmin, Descamps_2005 does not appear in refs.bib, I think. Please add it there or subtract the callout here, whatever is most appropriate.
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

%G3
To achieve goal G3 from the Introduction i.e., to explore, using the lottery model, the likelihood that species can coexist with
ATAs but not without, or vice versa, we plotted the IGR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
as well as the
IGR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[EC]}$. 
For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
we were interested in the values of $\mu_1-\mu_2$ for which one of these
quantities is positive and the other is negative; this set of values characterizes the species which can coexist
with ATAs, but not without, or vice versa. The one dimensional space of values $\mu$ is a species trait space,
and $\mu_1$ and $\mu_2$ can be regarded as two points in it. But we instead define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
Points of the trait-difference space characterize possible combinations of traits of the two competing species in our model. For fixed values of 
$\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of 
trait-difference space to be the set of 
values $\mu_1 - \mu_2$ for which the the IGR, $\olr_{i \bs i}$, is positive. This 
is the subspace for which the two species in the lottery model coexist in the presence of ATAs.
We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of values
$\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[EC]}$ is positive.
This is the subspace for which the two species in the lottery model coexist in the absence of 
ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ is intended to indicate the
absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
Similarly, we define ``ATA exclusion space'' to be the set difference 
$\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[EC]}$
against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
$\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.

%***DAN: Methods paras here for G4, if we accomplish G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we consider the coexistence of diatom 
species established by chemostat batch experiments \citep{Descamps_2005}.
\citep{Descamps_2005} used measurements of growth of \emph{Cyclotella pseudostelligera} 
and \emph{Fragilaria crotonensis} over a range of temperatures to parametrize a competitive 
resource model as follows:
put model here
%***DAN: Jasmin, please add the model. Hard to evaluate the text without it.
Via simulations, \cite{Descamps_2005} showed that coexistence of \emph{C. pseudostelligera} 
and \emph{F. crotonensis} occurs with oscillating temperature fluctuations, 
$\theta(t)=\theta_0 (t) +  a\sin(2\pi t/P)$, 
%***DAN: Jasmin, should \theta_0 depend on t? It should not, right? Please fix if I'm right.
defined by parameters $\theta_0 = 18, a = 6, P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down their coexistence into the contrirbuting mechanisms, 
and here, we consider varying parameters of temperature fluctuations and further decompose the 
coexistence of \emph{C. pseudostelligera} and \emph{F. crotonensis} to account for contributions 
of ATAs and ATA effects as described above. Our decomposition is defined by $E(t)=\theta(t)$, 
$C_j(t)=(K_j(t)+S(t))/S(t)$ and $r_j(E_j,C_j)=\frac{V_j(E)}{C_j} -D$. Thus in our exploration, 
environmental fluctuation is temperature fluctuation. We first considered the scenario where 
\emph{F. crotonensis} is the invader and emph{C. pseudostelligera} is the resident to compute 
the coexistence decomposition for \emph{F. crotonensis}. 
%put reverse in sup mat?
%***DAN: Yes, I think that would be a good idea. I forgot about that - we were able to just consider 1 invading 2 for 
%the lottery model, but cannot really do that here, right?
%do we need to explain how we separated the asymmetries using the sims??
%***DAN: I am not sure what you mean by this question, Jasmin
As for G2, we first plotted the terms of equation \ref{eq:full_decomp} for the chemostat 
model against independent parameters defining temperature oscillation, $\theta_0, a$, and 
$P$, a total of three plots. For each plot, we set two of the parameters constant at the 
original parameters which coexistence was experimentally found in \cite{Descamps_2005}, 
and let the remaining temperature parameter vary. 
We let $\theta_0$ vary in $[16,19]$, $a$ vary in $[1,6]$, and $P$ vary in $[20,120]$. 
Relative importance of contributions was judged from the 
plots to assess whether and to what extent ATA effects on coexistence were meaningfully 
large for the lottery model, compared to 
the contributions of other mechanisms. 
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).

@@

%***DAN: Jasmine, some changes are needed to the figure itself. Please make the text in the legend bigger.
%It looks as though a blue line is for ATA effects, and a red line is for the effects of correlation per se
%(that's what it says in the legend),
%but that differs from how you did it in the other figs. Or perhaps the legend is just wrong. Please check and
%fix so as to be consistent with the rest of the figures in which color represents which quantity and which
%notation represents which quantity. Also, please change panel labels to be of the format "(a)" instead of 
%"A".
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to species coexistence in the diatom system of \cite{Descamps_2005} 
is meaningfully large compared to 
other components of the MCT decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
%***DAN: This is for one species invading the other, right? Need to specify that, and say which is 
%invading which, and, at the end of the 
%caption indicate that the reverse invasion scenario is in sup mat and cite the corresponding figure there.
For each panel, 
two of the three parameters defining the temperature oscillation (i.e., the amplitude, $a$; period, $P$; 
and mean, $\theta_0$) are fixed at the original coexistence values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, \theta_0=18^\circ$C). 
%***DAN: I added units for a  and \theta_0 here, please check these are correct 
The IGR and components of the decomposition
are then plotted against the other parameter. The original value of the varying parameter and coexistence quantities at that value is indicated by the vertical dashed magenta lines. The contribution of ATAs (red line)
has magnitude comparable to other components and, in these parameters, is always positive, indicating that ATAs can contribute meaningfully
to diatom coexistence. In some instances, the IGR line (orange) falls 
between the ATA line (red)
and the $x$ axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the weaker competitor. Standard errors of plotted quantities ?.
%***DAN: Don't forget to fill in the standard error stuff
}  \label{fig:fig5}
\end{figure}

>>>>

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Descamps_2005, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

%G3
To achieve goal G3 from the Introduction, i.e., to explore, using the lottery model, the likelihood that species can coexist with
ATAs but not without, or vice versa, we plotted the IGR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
as well as the
IGR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[EC]}$. 
For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
we were interested in the values of $\mu_1-\mu_2$ for which one of these
quantities is positive and the other is negative; this set of values characterizes the species which can coexist
with ATAs, but not without, or vice versa. The one dimensional space of values $\mu$ is a species trait space,
and $\mu_1$ and $\mu_2$ can be regarded as two points in it. But we instead define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
Points of the trait-difference space characterize possible combinations of traits of the two competing species in our model. For fixed values of 
$\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of 
trait-difference space to be the set of 
values $\mu_1 - \mu_2$ for which the the IGR, $\olr_{i \bs i}$, is positive. This 
is the subspace for which the two species in the lottery model coexist in the presence of ATAs.
We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of values
$\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[EC]}$ is positive.
This is the subspace for which the two species in the lottery model coexist in the absence of 
ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ is intended to indicate the
absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
Similarly, we define ``ATA exclusion space'' to be the set difference 
$\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[EC]}$
against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
$\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.

%***DAN: Methods paras here for G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory system of freshwater diatoms
explored by \cite{Descamps_2005}.
\citep{Descamps_2005} used measurements of growth of \emph{Cyclotella pseudostelligera} 
and \emph{Fragilaria crotonensis} over a range of temperatures to parametrize a competitive 
resource model as follows:
put model here
%***DAN: Jasmin, please add the model. Hard to evaluate the text without it.
After using lab experiments to parameterize the model, \cite{Descamps_2005} use simulations to
show that coexistence of \emph{C. pseudostelligera} 
and \emph{F. crotonensis} occurs when temperature fluctuates periodically, 
$\theta(t)=\theta_0 (t) +  a\sin(2\pi t/P)$, 
%***DAN: Jasmin, should \theta_0 depend on t? It should not, right? Please fix if I'm right.
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
%***DAN: I added units. Please check.
\cite{Ellner_2016, Ellner_2019} broke down their coexistence into the contributing mechanisms, 
and here, we consider varying parameters of temperature fluctuations and further decompose the 
coexistence of \emph{C. pseudostelligera} and \emph{F. crotonensis} to account for contributions 
of ATAs and ATA effects as described above. Our decomposition is defined by $E(t)=\theta(t)$, 
$C_j(t)=(K_j(t)+S(t))/S(t)$ and $r_j(E_j,C_j)=\frac{V_j(E)}{C_j} -D$. Thus in our exploration, 
environmental fluctuation is temperature fluctuation. We first considered the scenario where 
\emph{F. crotonensis} is the invader and emph{C. pseudostelligera} is the resident to compute 
the coexistence decomposition for \emph{F. crotonensis}. 
%put reverse in sup mat?
%***DAN: Yes, I think that would be a good idea. I forgot about that - we were able to only consider 1 invading 2 for 
%the lottery model, because of some simplifying assumptions we were able to make, but cannot really do that here, right?
%do we need to explain how we separated the asymmetries using the sims??
%***DAN: I am not sure what you mean by this question, Jasmin
As for G2, we first plotted the terms of equation \ref{eq:full_decomp} for the chemostat 
model against independent parameters defining temperature oscillation, $\theta_0, a$, and 
$P$, a total of three plots. For each plot, we set two of the parameters constant at the 
original parameters which coexistence was experimentally found in \cite{Descamps_2005}, 
and let the remaining temperature parameter vary. 
We let $\theta_0$ vary in $[16,19]$, $a$ vary in $[1,6]$, and $P$ vary in $[20,120]$. 
Relative importance of contributions was judged from the 
plots to assess whether and to what extent ATA effects on coexistence were meaningfully 
large for the lottery model, compared to 
the contributions of other mechanisms. 
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).

@@

%***DAN: Jasmin, some changes are needed to the figure itself. Please make the text in the legend bigger.
%It looks as though a blue line is for ATA effects, and a red line is for the effects of correlation per se
%(that's what it says in the legend),
%but that differs from how you did it in the other figs. Or perhaps the legend is just wrong. Please check and
%fix so as to be consistent with the rest of the figures in which color represents which quantity and which
%notation represents which quantity. Also, please change panel labels to be of the format "(a)" instead of 
%"A". Make the panel labels bigger, as well as the axis numbers and labels. 
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to species coexistence in the diatom system of \cite{Descamps_2005} 
is meaningfully large compared to 
other components of the MCT decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
%***DAN: This is for one species invading the other, right? Need to specify that, and say which is 
%invading which, and, at the end of the 
%caption indicate that the reverse invasion scenario is in sup mat and cite the corresponding figure there.
For each panel, 
two of the three parameters defining the temperature oscillation (i.e., the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) are fixed at the original values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C) and 
%***DAN: I added units for a  and \theta_0 here, please check these are correct 
the IGR and components of the MCT decomposition
are then plotted against the other parameter. The original value of the varying parameter 
is indicated by the vertical dashed magenta lines. ATA contributions 
had magnitude comparable to other components, 
%***DAN: Make sure this is still true after the correct version of C is put in, and after we
%resolve the potential problem with the wiggles on the plot. Reword if not.
indicating that ATAs can contribute meaningfully
to diatom coexistence in this system. In some instances, the IGR line (orange) falls 
between the ATA line (red)
and the horizontal axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the weaker competitor. 
%***DAN: Not the weaker competitor, actually, right, but rather whichever competitor was being considered as the invader
%for these plots? If I'm right about that then please reword accordingly. 
Standard errors of plotted quantities ?.
%***DAN: Don't forget to fill in the standard error stuff
%***DAN: Here is where you add a statement citing another figure in the sup mat that takes the other
%species as the invader.
}  \label{fig:fig5}
\end{figure}
